---
title: "Iron Tolerance in Blood Feeding vs Obligate Non Biting Populations of the Pitcher Plant Mosquito, Wyeomyia smithii"  
subtitle: "Replication of Honors Thesis Analysis and Figures"
author: "Sophia Bevans"
output: github_document
---
```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = "~/UofO/Thesis/honors-thesis-analysis-mosquito")
```

```{r, include=FALSE}
source("external_analysis_code.R", local = knitr::knit_global())
```

Reproduction of analysis detailed in experimental design given by Dr. Bill Bradshaw and Dr. Christina Holzapfel in document "Sophia Experiments II".

## Data
Wyeomyia smithii populations: KC, ML (non-biting), WI, LI (biting) were collected from Maine, Wisconsin, Florida, and Alabama, respectively, in the summer of 2016.  
Experimental groups: 0 = treatment (water), 1 = 13mg/L ferrous sulfate, 2 = 130mg/L ferrous sulfate  
Column names refer to population and experimental group (ex: WI0 is control group from Florida)  

Treatment group ML0 is missing in the data due to issues with experimental equipment.

pupae contains date of pupation for pupae in each treatment group  
pupa_sex contains sex and eclosion date of pupae in each treatment group  
adult_sex contains sex and date of death for adults in each treatment group    
eggs contains date and counts of eggs from each treatment group   
hatch counts contains date and counts of hatch from each treatment group  


### Initial Data Format
```{r view data}
head(pupa_sex_orig)
```

### Pivoted Tables with Formatted Dates:
```{r}
#functions and code in external_analysis_code.R
head(pupa_sex)
```

### Summary Statistics
```{r pupae, eggs and hatch, message=FALSE}
head(pupae_sum)
```

```{r adult eclosion}
head(pupa_sex_sum)
```

### Mean Adult Longevity (Mean Eclosion Date - Mean Death Date)
```{r longevity}
head(longev)
```

## Nested ANOVA, Sex and Longevity in Treatment Group
```{r anova longevity}
longev_anova <- aov(longevity ~ Treatment / Biting_Propensity / Sex, data = longev)
summary(longev_anova)
```

### Mean Eggs and Hatched Eggs Per Eclosed Female, Arcsine Transformed Proportion of Hatched Eggs
```{r eggs and hatch}
head(eggs_hatch)
```

## Nested ANOVA, Fertility, Fecundity, Reproductive Success
### Fertility
```{r anova fertility (% eggs hatched)}
fertility_anova <- aov(arcsin_ph ~ Treatment / Biting_Propensity, data = eggs_hatch)
summary(fertility_anova)
```

### Fecundity
```{r anova fecundity (eggs per female)}
fecundity_anova <- aov(log(EggsPerFemale) ~ Treatment / Biting_Propensity, data = eggs_hatch)
summary(fecundity_anova)
```

### Lifetime Reproductive Success
```{r anova reproductive success (hatch per female)}
repsuccess_anova <- aov(log(HatchPerFemale) ~ Treatment / Biting_Propensity, data = eggs_hatch)
summary(repsuccess_anova)
```

## Figures

```{r longevity figure, warning= FALSE}
longev %>% 
  ggplot() + geom_point(aes(x = Treatment, y = longevity, shape = Sex, color = Biting_Propensity), size = 8, stroke = 1.25, position = position_dodge(width = 0.5)) + 
  labs(x = "Iron Treatment", y = "Mean Longevity (Days)", title =  "Mean Adult Longevity of Males and Females", caption = "Ferrous sulfate concentrations:\n0: 0mg/mL, 1: 13mg/mL, 2: 130 mg/mL.\nSignificance of nested ANOVA is given by: *P<0.05") + 
  theme_classic() + scale_shape_manual(values = c(1, 2)) + 
  scale_color_manual(values = c("red", "dodgerblue")) + ylim(0, 20) + 
  annotate("text", x = 0.75, y = 2, label = expression(atop(Fe^"*",
                                                    Biting^"*")))
```

```{r fecundity figure, warning=FALSE}
eggs_hatch %>% 
  ggplot() + geom_point(aes(x = Treatment, y = log10(EggsPerFemale), color = Biting_Propensity), shape = 1, size = 8, stroke = 1.25, position = position_dodge(width = 0.5)) + 
  labs(x = "Iron Treatment", y = "Log(Eggs per Female)", title =  "Lifetime Fecundity", subtitle = "Eggs Per Eclosed Females", caption = "Ferrous sulfate concentrations:\n0: 0mg/mL, 1: 13mg/mL, 2: 130 mg/mL.\nSignificance of nested ANOVA is given by: *P<0.05") + 
  theme_classic() + 
  scale_color_manual(values = c("red", "dodgerblue")) + ylim(0, 2) + 
  annotate("text", x = 0.75, y = 0.25, label = expression(Biting^"*"))
```

```{r reproductive success figure, warning=FALSE}
eggs_hatch %>% 
  ggplot() + geom_point(aes(x = Treatment, y = log10(HatchPerFemale), color = Biting_Propensity), shape = 1, size = 8, stroke = 1.25, position = position_dodge(width = 0.5)) + 
  labs(x = "Iron Treatment", y = "Log(Hatch per Female)", title =  "Lifetime Reproductive Success", subtitle = "Hatch Per Eclosed Females", caption = "Ferrous sulfate concentrations:\n0: 0mg/mL, 1: 13mg/mL, 2: 130 mg/mL.\nSignificance of nested ANOVA is given by: *P<0.05") + 
  theme_classic() + 
  scale_color_manual(values = c("red", "dodgerblue")) + ylim(0, 2) + 
  annotate("text", x = 0.75, y = 0.25, label = expression(Biting^"*"))
```

```{r fertility figure}
eggs_hatch %>% 
  ggplot() + geom_point(aes(x = Treatment, y = arcsin_ph, color = Biting_Propensity), shape = 1, size = 8, stroke = 1.25, position = position_dodge(width = 0.5)) + 
  labs(x = "Iron Treatment", y = "Arcsine Transformed Frequency Hatched", title =  "Fertility", subtitle = "Frequency of Eggs Hatched", caption = "Ferrous sulfate concentrations:\n0: 0mg/mL, 1: 13mg/mL, 2: 130 mg/mL.") + 
  theme_classic() + 
  scale_color_manual(values = c("red", "dodgerblue")) + ylim(0, 2)
```

